PLANETESIMAL ACCRETION AT SHORT ORBITAL PERIODS

### 2022-10-12 main.tex
Edit summary + discussion

In this work, we have demonstrated that planetary embryo formation
operates in two distinct modes in a planet-forming disk. In the first
mode, gravitational feedback from the growing embryos heats the
remaining planetesimals and results in a dynamically cold population
of embryos with a significant amount of residual planetesimals. This
corresponds to the ``oligarchic growth'' case revealed by \citep{kokubo98}. In the second mode, the gravitational feedback does not operate, embryos quickly sweep up all planetesimals, and grow about twice as large as is predicted by oligarchic growth. The distinction between these modes is determined by ratio between the physical size of the bodies and their Hill radius.

We have demonstrated the outcome of both accretion modes through a
sparsely sampled parameter study. The initial planetesimal distribution can be described in terms of two dimensionless constants, $\alpha$ and $\beta$, which describe the ratio between the physical radius of the planetesimals and the Hill ($r_{h}$) and gravitational ($r_{g}$) radii, respectively. For a fixed planetesimal mass and radius, $\alpha$ scales with the orbital period and $\beta$ scales with the level of dynamical excitation of the disk. We showed that $\alpha \ll 1$ leads to oligarchic growth, while a non-negligible $\alpha$ produces this new non-oligarchic growth mode. We find that the resulting mass distribution, along with the final eccentricities of the embryos and residual planetesimals, ends up the same, regardless of the initial value of $\beta$.

So long as the density of the bodies does not significantly change as
their mass distribution evolves, this ratio, and therefore the
boundary between these accretion modes is set entirely by the distance
from the star. Because both the physical and Hill radii of the bodies
grows as $M^{1/3}$, the boundary between growth remains at a fixed
location in the disk during the planetesimal accretion process.

We have verified this fact by testing the outcome of the planetesimal
accretion process for a variety of solid surface density
profiles. Although altering the surface density does affect the
resulting masses of the embryos, the location of the boundary
separating the growth modes is remarkably similar among all of our simulations. We have verified this by comparing the resulting embryo masses to the isolation mass, in addition to highlighting qualitative differences in the accretion history of embryos on both sides of the boundary.

Lastly, we quantified the way in which our use of perfect accretion and an inflated collision cross section, both meant to make the simulations more computationally tractable, affect the non-oligarchic growth mode and the location of the accretion boundary. We showed that although the use of perfect accretion speeds up the growth of the embryos, it does not affect the masses or orbital properties of the resulting bodies. Because the inflated collision cross sections artificially reduce the density of the planetesimals, this shortcut does shift the accretion boundary outward. However, it moves in a predictable fashion and we verified this by running an extra simulation with a modestly smaller collision cross section. In a real planet-forming disk, one would expect this boundary to lie somewhere between 2 and 10 days, depending on the composition of the planetesimals.

To date, there have been no other studies of planetesimal accretion
with such a large value of $\alpha$. However, a value of $\alpha = 1$
corresponds to the Roche limit of a three-body system, and so one
might wonder this high-$\alpha$ accretion mode might be relevant for a
circumplanetary system. There is a small collection of previous works
which use N-body methods to examine in-situ satellitesimal accretion
\citep{ida97, richardson00, ida20}. \citet{ida97} was able to form 1-2
large moons just exterior to the Roche limit, depending on the extent of the disk with very little satellitesimal material left over. The widest disk they modeled extended out to $\alpha = 0.5$. Qualitatively, this result is very similar to the short period planetesimal accretion mode observed in our simulations. \citet{ida20} modeled a much wider satellitesimal disk, which extends out to about $\alpha \approx 0.05$. Inside to the $\alpha = 0.1$ accretion boundary (which lies near $15 R_{U}$ in figure 1 of \citet{ida20}), bodies grow beyond the isolation mass, while the opposite is true on the other side of the boundary. In addition, a residual population of satellitesimals is still present beyond the boundary, which suggests that oligarchic growth is indeed operating on the far side.

Presently, the implications that this non-oligarchic accretion mode has for the formation of short-period terrestrial planets, and whether the accretion boundary would leave any lasting imprint on the final orbital architecture, is unclear. One point that our results do highlight is that the initial conditions used for most late stage planet formation simulations are overly simplistic. \citet{clement20} recently simulated planetesimal accretion in a disk extending from the orbit of Mercury to the asteroid belt and found that the disk never reaches a state in which equally-spaced, isolation mass embryos are present everywhere simultaneously. Instead, different annuli reach a `giant impact' phase at different times, preventing the onset of a global instability throughout the entire disk, as is common in classic terrestrial planet formation models \citep{chambers01, raymond09}.

% Any examples of studies of STIPs that use these simplistic ICs? Could the accretion boundary help reconcile differences with kepler systems?

To connect these accretion modes to the final orbital architecture,
and to ultimately determine what implications an in situ formation model has for the growth of STIPs, we will continue to evolve the final simulation snaphots presented here with a hybrid symplectic integrator. The final distribution of planets formed, along with composition predictions generated by applying cosmochemical models to our initial planetesimal distributions and propagating compositions through the merger trees, will be examined in a follow-up paper.

>>>>

In this work, we have demonstrated that planetary embryo growth
can operate in two distinct modes in a planet-forming disk. In the first
mode, gravitational feedback from the growing embryos heats the
remaining planetesimals and results in a dynamically cold population
of embryos with a modest amount of residual planetesimals. This
corresponds to the ``oligarchic growth'' case revealed by \citep{kokubo98}, which is often used as a starting point for late-stage accretion models (e.g. \citet{kokubo02, raymond05, raymond06}). In the second mode, the gravitational feedback does not operate, embryos quickly sweep up all planetesimals, and grow larger and less uniformly spaced than those produced by oligarchic growth. The distinction between these modes is determined by ratio between the physical size of the bodies and their Hill radius, which sets the relative contribution of close gravitational encounters and collisions to the dynamical evolution of the disk.

We have demonstrated the outcome of both accretion modes through a
sparsely sampled parameter study. The initial planetesimal distribution can be described in terms of two dimensionless constants, $\alpha$ and $\beta$, which describe the ratio between the physical radius of the planetesimals and the Hill ($r_{h}$) and gravitational ($r_{g}$) radius, respectively. For a fixed planetesimal composition, $\alpha$ scales with the orbital period and $\beta$ scales with the level of dynamical excitation of the disk. We showed that $\alpha \ll 1$ leads to oligarchic growth, while a non-negligible $\alpha$ produces this newly revealed non-oligarchic growth mode. Within this non-oligarchic mode, we find that the resulting masses and eccentricities of the embryos come out the same, regardless of the initial value of $\beta$.

So long as the density of the bodies does not significantly change as
their mass distribution evolves, this ratio, and therefore the
boundary between these accretion modes is set entirely by the distance
from the star. Because both the physical and Hill radii of the bodies
grows as $M^{1/3}$, the growth mode boundary remains stationary
 in the disk during the planetesimal accretion process.

We have verified that the growth boundary location is independent of the
solid surface density distribution by testing the outcome of the planetesimal
accretion process for a variety of solid
profiles. Although altering the surface density does affect the
resulting masses of the embryos, the location of the boundary
separating the growth modes is remarkably similar among all of our simulations.
In addition, the sizes of the feeding zones, along with qualitative differences in the
accretion history of embryos on both sides of the boundary provide further evidence to
suggest that oligarchy is not operating in the inner disk.

Finally, we examined the way in which our simplifying assumption of perfect accretion, along with the enhancement of the 
collision cross sections to speed up the simulations might alter the emergence and location of the non-oligarchic growth region. 
We showed that a much more restrictive collision model, in which only low-velocity collisions produce a merger, still allows for 
this growth mode to occur at the innermost part of the disk, where encounter speeds are most vigorous. In a real planet-forming 
disk, partial accretion events should allow growth to happen more quickly than what was seen in this test case, so this growth 
mode should certainly still occur. We also showed that the collision cross section enhancement moves the accretion boundary 
outward. We verified this by deriving a scaling relation between the boundary location and the bulk density of the planetesimals, 
and showing that the boundary moves to the predicted location when running a simulation with a slightly smaller inflation factor. 
For planetesimals with a realistic bulk density, this boundary should lie around 5 days in orbital period.

To date, there have been no other studies of planetesimal accretion
with such a large value of $\alpha$. However, a value of $\alpha = 1$
corresponds to the Roche limit of a three-body system, and so one
might wonder this high-$\alpha$ accretion mode might be relevant for a
circumplanetary system. There is a small collection of previous works
which use N-body methods to examine in-situ satellitesimal accretion
\citep{ida97, richardson00, kokubo00b, ida20}, although some of these simulations involve a complex interaction between spiral density waves formed inside of the Roche limit and the material exterior to it. \citet{ida97} was able to form 1-2
large moons just exterior to the Roche limit, depending on the extent of the disk with very little satellitesimal material left over. The widest disk they modeled extended out to $\alpha = 0.5$. Qualitatively, this result is very similar to the short period planetesimal accretion mode observed in our simulations. \citet{ida20} modeled a much wider satellitesimal disk, which extends out to about $\alpha \approx 0.05$. Inside to the $\alpha = 0.1$ accretion boundary (which lies near $15 R_{U}$ in figure 1 of \citet{ida20}), bodies grow beyond the isolation mass, while the opposite is true on the other side of the boundary. In addition, a residual population of satellitesimals is still present beyond the boundary, which suggests that oligarchic growth is indeed operating only on the far side.

Presently, the implications that this non-oligarchic accretion mode has for the formation of short-period terrestrial planets, and 
whether the accretion boundary would leave any lasting imprint on the final orbital architecture, is unclear. One point that our 
results do highlight is that the initial conditions used for most late-stage planet formation simulations are overly simplistic. 
\citet{clement20} recently simulated planetesimal accretion in a disk extending from the orbit of Mercury to the asteroid belt and 
found that the disk never reaches a state in which equally-spaced, isolation mass embryos are present everywhere 
simultaneously. Instead, different annuli reach a `giant impact' phase at different times, preventing the onset of a global instability 
throughout the entire disk, as is common in classic terrestrial planet formation models \citep{chambers01, raymond09}.

To connect these accretion modes to the final orbital architecture,
and to ultimately determine what implications an in-situ formation model has for the growth of STIPs, we will evolve the final 
simulation snapshots presented here with a hybrid-symplectic integrator for Myr timescales. The final distribution of planets 
formed, along with composition predictions generated by applying cosmochemical models to our initial planetesimal distributions 
and propagating compositions through the collision trees, will be examined in a follow-up paper.
